Weighted Model Counting with Conditional Probabilities

### 2020-07-05 paper.tex
Non-Boolean Bayesian networks: version 1

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{complexity}
\usepackage{soul} % temp

@@

\item For both DQMR and Plan Recognition, I will need two DNE $\to$ CNF
  encoders. I should formalise the algorithms behind them as well, since they
  are described informally.

@@

\item Along with comparing the (3?) encodings, I could also compare this
  to using ADDs directly on Bayesian networks.

@@

  probabilities.

@@

\paragraph{Notes.}
\begin{itemize}
\item Thesis: many important computational problems are solved by encoding them
  as WMC. But if the weight function is extended to `conditional probabilities',
  the problem becomes easier because current approaches need workarounds in
  order to be independent.
\item Shorter thesis: many important problems are encoded as WMC, but they can
  be encoded in a better way if we allow for conditional probabilities.
\item We extend the Gaifman graph to add edges when two variables occur in the
  same CPT (e.g., including the edge from $A$ to $B$ when the CPT is $\Pr(A \mid
  B)$).
\item Extra benefit: one does not need to come up with a way to turn some probability
distribution to into a fully independent one.
\item Trivial CPTs such as $\Pr(A \mid B) = \Pr(\neg A \mid B) = 0.5$, the ADDs
  of which simplify to a single number, are put in the first cluster.
\item Observation: by inspecting the BN, we could identify CPTs that could be
  completely ignored, but maybe good heuristics take care of that anyway.
\item Observation: ADDs have a restrict command. I'm not going to use it, but it
  could be nice.
\item Alternative scenario to consider for future work: \#\SAT{} solver
  generates solutions which are then used to restrict the ADDs that calculate
  probabilities.
\item Important future work: replacing ADDs with
  AADDs\footnote{\url{https://github.com/ssanner/dd-inference}} is likely to
  bring performance benefits. Other extensions:
  \begin{itemize}
  \item FOADDs can represent first order statements;
  \item XADDs can replace WMI for continuous variables;
  \item ADDs with intervals can do approximations.
  \end{itemize}
\item In preliminary results, the SBK05 encoding gives the wrong asnwer in 60
  QMR instances (all with evidence files). Sometimes that answer is $\gg 1$.
\item Similarly, cd06 is supposed to be the fastest, but seems to be one of the
  slowest (along with cd05). Perhaps the best encoding for arithmetic circuits
  has very little to do with the best encoding for CNF WMC solvers?
\item When the Bayesian network has an evidence file, we compute the probability
  of evidence. Otherwise, we pick the last (i.e., last to be mentioned) node in
  the Bayesian network to be the goal. This applies to the Grid data set (as
  intended) and also to two instances of Plan Reconstruction and roughly half of
  2004-PGM.
\end{itemize}

@@

\section{How to represent a Bayesian network}

Let $G = (V, E)$ be a DAG, and let $(X_v)_{v \in V}$ be the discrete random
variables (RVs). Let $(d_v)_{v \in V}$ denote the numbers of values that each RV
can take and assume that $d_v$ is a power of two for all $v \in V$.
\begin{enumerate}
\item For each $v \in V$, let $A_v$ be a set of $\log_2 d_v$ Boolean variables.
  (If $d_v$ is not a power of two, we would need to manually forbid each illegal
  value, e.g., by adding formulas such as $\neg(x_1 \land x_2 \land x_3 \land
  \neg x_4)$ to the theory. Alternatively, we could set their probability to
  zero, but the former should be better.)
\item For $v \in V$:
  \begin{enumerate}
  \item If $v$ has no parents, we can directly provide a probability measure on
    $\mathcal{F}(A_v)$.
  \item Otherwise, we need a probability measure on
    \[
      \mathcal{F} \left( A_v \cup \bigcup_{(u, v) \in E} A_u \right).
    \]
    For any element $a_u \land a_v$ in this free Boolean algebra such that $a_v
    \in A_v \setminus \{ 0, 1 \}$, and $a_u \not\in A_v$, we can set
    \[
      m(a_u \land a_v) = \Pr(a_v \mid a_u)\Pr(a_u).
    \]
  \end{enumerate}
\item When a \#\SAT{} algorithm asks for the weight of a model $\bigwedge_{v \in
    V} l_v$, split it into segments that correspond to each leaf, take the
  probabilities from each such node, and multiply them. NB: The segments will
  intersect, e.g., with
  \[
    \begin{tikzpicture}
      \node[latent] (C) {$C$};
      \node[latent, below left = of C] (A) {$A$};
      \node[latent, below right = of C] (B) {$B$};
      \node[latent, right = 2 of C] (D) {$D$};
      \edge {C} {A, B};
    \end{tikzpicture}
  \]
  $a \land b \land c \land d$ will be split into $(a \land c) \land (b \land c)
  \land d$.
\end{enumerate}

\subsection{Bayesian networks as Boolean algebras}

\begin{table}
  \caption{Examples of how Bayesian networks encode as Boolean algebras. Here,
    $A$ denotes $\mathcal{F}\{a\}$ and $AB$ denotes $\mathcal{F}\{a, b\}$.}
  \centering
  \begin{tabular}{cc}
    \toprule
    Bayesian network & Boolean algebra \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={(A.base)}]
      \node[latent] (A) {$A$};
    \end{tikzpicture} & $A$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={(A.base)}]
      \node[latent] (A) {$A$};
      \node[latent,right =of A] (B) {$B$};
    \end{tikzpicture} & $A + B$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={($(A.base)!.5!(B.base)$)}]
      \node[latent] (A) {$A$};
      \node[latent,below =of A] (B) {$B$};
      \edge {A} {B};
    \end{tikzpicture} & $AB$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={($(A.base)!.5!(B.base)$)}]
      \node[latent] (A) {$A$};
      \node[latent,below =of A] (B) {$B$};
      \node[latent,right =of A] (C) {$C$};
      \edge {A} {B};
    \end{tikzpicture} & $B +_A C$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={($(A.base)!.5!(B.base)$)}]
      \node[latent] (A) {$A$};
      \node[latent,below =of A] (B) {$B$};
      \node[latent,right =of A] (C) {$C$};
      \node[latent,below =of C] (D) {$D$};
      \edge {A} {B};
      \edge {C} {D};
    \end{tikzpicture} & $B +_{A+C} D$ \\
    \midrule \\[-1.3em]
    \begin{tikzpicture}[baseline={($(A.base)!.5!(B.base)$)}]
      \node[latent] (A) {$A$};
      \node[latent,below left =of A] (B) {$B$};
      \node[latent,below right=of A] (C) {$C$};
      \node[latent,right = 2 of A] (D) {$D$};
      \node[latent,below =of D] (E) {$E$};
      \edge {A} {B,C};
      \edge {D} {E};
    \end{tikzpicture} & $B +_{A+D} C +_{A+D} E$ \\
    \bottomrule
  \end{tabular}
\end{table}

To turn a Bayesian network into a Boolean algebra, run $f(\{ v \in V \mid
\mathrm{outdeg}(v) = 0 \})$, where $f$ works as follows on an arbitrary $A
\subseteq V$:
\begin{enumerate}
\item $S \gets \{ u \mid (u, v) \in E\text{, } v \in A \}$ (i.e., $S$ is the set
  of parents of the elements in $A$).
\item If $S = \emptyset$, return $\sum A$ (a coproduct).
\item Otherwise, return $\sum_{f(S)} A$ (a pushout).
\end{enumerate}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{complexity}
\usepackage{soul}

@@

@@

@@

  probabilities. Ditto for non-binary values (the shorter version).

@@

\paragraph{Notes.}
\begin{itemize}
\item Thesis: many important computational problems are solved by encoding them
  as WMC. But if the weight function is extended to `conditional probabilities',
  the problem becomes easier because current approaches need workarounds in
  order to be independent.
\item Shorter thesis: many important problems are encoded as WMC, but they can
  be encoded in a better way if we allow for conditional probabilities.
\item We extend the Gaifman graph to add edges when two variables occur in the
  same CPT (e.g., including the edge from $A$ to $B$ when the CPT is $\Pr(A \mid
  B)$).
\item Extra benefit: one does not need to come up with a way to turn some probability
distribution to into a fully independent one.
\item Trivial CPTs such as $\Pr(A \mid B) = \Pr(\neg A \mid B) = 0.5$, the ADDs
  of which simplify to a single number, are put in the first cluster.
\item Observation: by inspecting the BN, we could identify CPTs that could be
  completely ignored, but maybe good heuristics take care of that anyway.
\item Observation: ADDs have a restrict command. I'm not going to use it, but it
  could be nice.
\item Alternative scenario to consider for future work: \#\SAT{} solver
  generates solutions which are then used to restrict the ADDs that calculate
  probabilities.
\item Important future work: replacing ADDs with
  AADDs\footnote{\url{https://github.com/ssanner/dd-inference}} is likely to
  bring performance benefits. Other extensions:
  \begin{itemize}
  \item FOADDs can represent first order statements;
  \item XADDs can replace WMI for continuous variables;
  \item ADDs with intervals can do approximations.
  \end{itemize}
\item In preliminary results, the SBK05 encoding gives the wrong asnwer in 60
  QMR instances (all with evidence files). Sometimes that answer is $\gg 1$.
\item Similarly, cd06 is supposed to be the fastest, but seems to be one of the
  slowest (along with cd05). Perhaps the best encoding for arithmetic circuits
  has very little to do with the best encoding for CNF WMC solvers?
\item When the Bayesian network has an evidence file, we compute the probability
  of evidence. Otherwise, we pick the last (i.e., last to be mentioned) node in
  the Bayesian network to be the goal (the positive value for binary networks
  and the last value in the general case). This applies to the Grid data set (as
  intended) and also to two instances of Plan Reconstruction and roughly half of
  2004-PGM.
\item Bayesian networks are often solved in a compile once, query many times
  fashion. This can be achieved using ADDMC by selecting a subset $S$ of
  variable we may want to query over and running ADDMC while excluding $S$ from
  variable elimination/projection/$\exists$.
\end{itemize}

@@

\section{Representation of conditional probability tables}

The probability of a variable with $n$ values $x_1, \dots, x_n$ (whether
conditioned on something or not) can be represented with an ADD of size
$\Theta(n^2)$ by representing each of $\Pr(x_1), \Pr(x_2 \mid x_1), \dots,
\Pr(x_n \mid x_1, \dots, x_{n-1})$ by a separate ADD. For example, if $n=3$, the
last of the three ADDs would look like this
\[
  \begin{tikzpicture}
    \node[ellipse,draw] (x3) {$x_3$};
    \node[ellipse,draw,below left = of x3] (x21) {$x_2$};
    \node[ellipse,draw,below right = of x3] (x22) {$x_2$};
    \node[ellipse,draw,below = of x21] (x11) {$x_1$};
    \node[ellipse,draw,below = of x22] (x12) {$x_1$};
    \node[rectangle,draw,below = of x11] (n2) {$0.7$};
    \node[rectangle,draw,left = of n2] (n1) {$0$};
    \node[rectangle,draw,below = of x12] (n3) {$0.3$};
    \node[rectangle,draw,right = of n3] (n4) {$1$};

    \draw (x3) edge (x21);
    \draw (x21) edge (n1);
    \draw (x11) edge (n1);
    \draw[dashed] (x21) edge (x11);
    \draw[dashed] (x11) edge (n2);

    \draw[dashed] (x3) edge (x22);
    \draw[dashed] (x22) edge (x12);
    \draw[dashed] (x12) edge (n3);
    \draw (x22) edge (n4);
    \draw (x12) edge (n4);
  \end{tikzpicture}
\]
Such an ADD for $x_n$ has $2n+3$ nodes and $4n-2$ edges. The same information
can be represented in textual form as
\begin{align*}
  &\texttt{w 3 1 0} \\
  &\texttt{w 3 2 0} \\
  &\texttt{w 3 -1 -2 0.7}
\end{align*}
